---
title: "DGE Analysis"
author: "George W Kagugube"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background to the project

Manganese (Mn) is a trace element with multiple physiological function mainly 
through emzymatic regulation. for example, Mn is a cofactor of superoxide
dismutase 1 which is mitochondrial based enzyme critical to the reactive oxygen
species control in the mitochondria. 

Mn homeostasis is therefore important for normal physiological function. 

In this project, I sought to undestand the changes in the transcriptional profile 
of a Mn overloaded zebrafish. To understand the extent of the changes, I exposed
slc39a14<sup>+/+</sup> and slc39a14<sup>-/-</sub> to 20 µM MnCl<sub>2</sub>. 

I then analysed the datasets for differential gene expression (DGE). 

```{r echo=FALSE}
## Load and source the files needed for the analysis
source("/Users/gwk/Desktop/Bioinformatics/bulk-rnaseq-zebrafish-mouse-human/scripts/functionsneededforanalysis.R")
```

```{r}
## Read the mapped and quantified files into r for further analysis here 
file1 <- "1_S1_ReadsPerGene.out.tab"
dr <- '/Users/gwk/Desktop/PhD/Data/PhD_data/Brain/GZ11_star_output/star'

# This is output from star - these are files with ReadsPerGene.out.tab
countMatrix <- staroutput_preprocessing(dr, file1)

# If no sample information exists, you can create to match the countMatrix
samples <- read.csv("sample_information.csv", row.names = 1)
```

## Explore the datsets that have been imported
The files from STAR mapping are imported here. The code is part of a pipeline, 
that maps the reads to a reference genome (Gz11) using the two pass approach. 

To check for the structure of these two file i.e. the count files and sample info
documents allows us to make sure that we set up the files to be acceptade by 
DESeq2

```{r}
head(countMatrix)
head(samples, 15)
```

```{r}
## Fix the column names in the count matrix to match the sample information rownames
row_col_names <- as.vector(paste0("sample",rownames(samples)))
rownames(samples) <- row_col_names
colnames(countMatrix) <- row_col_names

## Makesure that sample rownames match the countmatrix colnames here
# The lines below must return true for all
# if false, please investigate the rownames match the countMatrix colnames
all(row.names(samples) %in% colnames(countMatrix))
all(row.names(samples) == colnames(countMatrix))
```

The experimental setup was such that WT and mutants were reared seperately. 
Therefore, there are a few aproaches to dealing with the variation that comes
with such a design. Analyse data within a genotype seperately, combine the data
and batch correct for the seperate starting point of the groups. Here I have tried 
all and compared the outputs from each. The most reliable comparision is within
genotype.

```{r}
# Split the data so that mutant and wt 
# Mutant sample information
mut <- samples %>%
  filter(Genotype == "mutant") %>%
  dplyr::select(Group)

mut

mutt <- c(3,4,8,9,10,11)
mutt

# Wild type sample information
wt <- samples %>%
  filter(Genotype == "wt") %>%
  dplyr::select(Group) 
wt
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
